Tracing the Meta-Level: PyPy's JIT Compiler

### 2009-04-01 talk/icooolps2009/paper.tex
Typos, and a note about the introduction.

Tracing JITs are an idea explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{mason_chang_efficient_2007}. The technique is now used by both and are now
being used by both Mozilla's TraceMonkey JavaScript VM \cite{XXX} and Adobe's
Tamarin ActionScript VM \cite{XXX}.

@@

\anto{
Such a history is called a \emph{trace}: it is a sequential list of
operations, togheter with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by emitting
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code: to ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example \texttt{if}s or indirect/virtual
calls.  When emitting the machine code, we turn every guard into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit from the machine code and continue the execution in other
ways.  
}

During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that is being interpreted at the moment anyway.

>>>>

\arigo{We should not start from scratch and insert as little details what
differs in our approach when compared to others; instead we should give a
higher-level overview and then focus on these details, and a couple of
references for more info about the "common" part.
%
In general there are many things that are never said at all.
I think the introduction should really be written from the point of view of
someone that has read already some papers for JavaScript.}

Tracing JITs are an idea explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{mason_chang_efficient_2007}. The technique is now
being used by both Mozilla's TraceMonkey JavaScript VM \cite{XXX} and Adobe's
Tamarin ActionScript VM \cite{XXX}.

@@

\anto{
Such a history is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by emitting
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code: to ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example \texttt{if}s or indirect/virtual
calls.  When emitting the machine code, we turn every guard into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit from the machine code and continue the execution in other
ways.  
}

During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that was being interpreted so far.
